The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-11 paper/iphas-dr2.tex
Update iphas-dr2.tex
The calibration section has now been gone through (catalogue next)

Unlike the broadbands, 
standard-star photometry is not available in the literature 
for the H$\alpha$ passband
and hence there is no formally recognised flux scale 
for it.
We can specify here, however, 
that the integrated in-band energy flux for Vega 
in the IPHAS \ha\ filter 
is $1.52 \times 10^{-7}$ ergs\,cm$^{-2}$\,s$^{-1}$ 
at the top of the Earth's atmosphere,
which is the flux obtained by folding 
Vega's SED with the filter transmission curve 
(uncorrected for the atmosphere and detector quantum efficiency,
which would otherwise scale down the narrowband flux by 0.707).
This is 3.14 magnitudes less than the flux captured 
by the much broader $r$ band
which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}
Note that this implies that the in-band flux corresponding to
zero magnitude is $1.56 \time 10^{-7}$ ergs\,cm$^{-2}$\,s$^{-1}$,
when the \ha\ magnitude for Vega is set by convention to 0.03.

@@

Unfortunately these schemes cannot be applied
directly to IPHAS
for two reasons. 
Firstly, the survey was carried out 
in competitive observing time
on a non-dedicated telescope, 
rendering the 2MASS approach 
of observing standards at a very high frequency
prohibitively expensive
-- in part because standard fields 
are very scarce within the Galactic Plane.
Secondly, the aim of IPHAS is to obtain magnitudes at a single epoch,
and hence stars are not normally observed at more than one epoch,
unless they happen to fall within a narrow overlap region 
between two neighbouring fields.

Although the IPHAS data does contain a significant number of 
such inter-field repeat measurements,
we have found the information contained
in these regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is prone to systematics at the level of 1--2\%.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is most strongly affected by a radial geometric distortion in the WFC,
which causes the pixel scale towards the edges 
to increase \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can combine to cause artificial zeropoint gradients 
across the survey
when they are used to constrain a global calibration
without other external constraints.

For these reasons, we have decided not to depend
on an ubercalibration-type scheme alone,
and have opted to involve an external reference survey
-- where available --
to bring the majority of our data onto a uniform calibration.
This is explained next.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring the vast majority of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the provisional calibration of IPHAS,
for which we estimate the $1\sigma$-uncertainty to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

With the aim of bringing IPHAS to a similar accuracy of $\sim$3\%,
we used the APASS catalogue to identify and adjust the calibration of all IPHAS fields 
which showed a magnitude offset larger than 3\% against APASS.
For this purpose,
the $r$- and $i$-band detection tables of each IPHAS field
were cross-matched against the APASS DR7 catalogue 
using a maximum matching distance of 1~arcsec.
The magnitude range was limited to
$13<r_{\rm APASS}<16.5$ and $12.5<i_{\rm APASS}<16.0$
in order to avoid sources 
brighter than the IPHAS saturation limit on one hand, 
and to avoid sources near the faint detection limit of APASS 
on the other hand.

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the $r$ and $i$ filters 
are very similar in both surveys.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses magnitudes relative to Vega.
Separate transformations were derived for sight-lines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
The variations at these different
sight-lines were found to be insignificant.
This is not surprising because heavily reddened objects 
are naturally scarce at $r<16$.

Having transformed APASS magnitudes into the IPHAS system,
we then computed the median magnitude offset 
for each field which contained at least 30 cross-matched stars.
This was the case for 48\% of our fields.
The mean offset was found to be
$0.014\pm0.104$~mag in $r$ and $0.007\pm0.108$~mag in $i$
(Table~\ref{tbl:offsets_before}).
A total of 4596 fields showed a median offset
exceeding $\pm$0.03~mag in either $r$ or $i$.

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of the fields in our data release.
Fortunately, these fields are predominantly located 
in the early part of the Galactic Plane (Fig.~\ref{fig:apass_r}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme, explained below,
which minimised the magnitude offsets between stars
located in the overlap regions with neighbouring fields.
Although we explained earlier that these overlap regions
are prone to systematics, the use of APASS enabled us to
keep the zeropoints of roughly half of the fields fixed,
which avoids these systematics from combining to introduce
artificial gradients across the survey.
In the next section we will show this to be true by validating
our calibration against SDSS.

A general solution to the problem of minimising the magnitude
differences between overlapping frames has previously been
described by~\citet{Glazebrook1994}.
In brief, there are two fundamental quantities to be
minimised between each pair of overlapping exposures, 
denoted by the indices $i$ and $j$. 
Firstly, the mean magnitude difference between stars in the overlap
region $\Delta_{ij}=\langle m_i-m_j\rangle=-\Delta_{ji}$ is a local
constraint. 
Secondly, to ensure the solution does not stray far 
from the existing calibration, 
the difference in zeropoints 
$\Delta\mathrm{ZP}_{ij}=-\Delta\mathrm{ZP}_{ji}$ 
between each pair of exposures must also be minimised.

@@

As explained above, we enforce a strong external constraint
on the solution by keeping the zeropoint fixed 
for the 6\,756 fields which have been compared
and calibrated against APASS earlier.
We hereafter refer to these fields as \emph{anchors}.
It is asserted that the zeropoints $a_i$ of these anchor fields 
are known and not solved for,
though they do appear in the vector $b_j$ as constraints.
In addition to the APASS-based anchors, 
we selected 3\,273 additional anchor fields by hand
to provide additional constraints in regions not covered by APASS.
These extra anchors were deemed to have reliable zeropoints 
based on 
(i) the information contained in the observing logs,
(ii) the stability of the standard star zeropoints during the night, and
(iii) photometricity statistics provided by the Carlsberg Meridian Telescope,
which is located $\sim$500~m from the INT.

@@

We then turned to the uniform calibration of the \ha\ data.
It is not possible to re-calibrate the narrowband 
in the same way as the broadbands,
because the APASS survey does not offer \ha\ photometry.
We can reasonably assume however,
that the corrections required for $r$ and \ha\ are identical
because the IPHAS data-taking pattern ensured 
that a field's \ha\ and $r$-band exposures
were taken at essentially the same time, 
separated only by the $\sim$30~s read-out time required by the WFC.
We have hence corrected the \ha\ zeropoints 
by re-using the zeropoint adjustments that were derived for the $r$ band
in the earlier steps.
An exception was made for 3\,101 fields
for which our quality-control routines revealed
zeropoint variations which exceeded 0.03~mag 
between consecutive fields,
suggesting strongly non-photometric conditions.
For good practice, the \ha\ zeropoints of these fields
were adjusted by solving Eqn.~\ref{eqn:matrix}
rather than linking them directly to the $r$-band shift.

@@

Having re-calibrated all fields to the expected APASS-accuracy of 3\%,
we then used a different survey to validate the results.
We have been able to exploit SDSS Data Release 9 \citep{Ahn2012}
for this purpose.
SDSS DR9 provides several strips at low
Galactic latitudes,
providing data across 18\% of the fields in our data release.
We cross-matched the IPHAS fields against the subset of
objects marked as reliable stars in the SDSS catalogue\footnote{
We used the CasJobs facility located at http://skyserver.sdss3.org/CasJobs
to obtain photometry from the SDSS {\sc photoprimary} table 
with criteria {\sc type = star}, {\sc clean = 1} and {\sc score $>$ 0.7}.}
in the same way as we did for APASS earlier,
with the exception of using fainter magnitude ranges of 
$15<r_{\rm SDSS}<18.0$ and $14.5<i_{\rm SDSS}<17.5$.
This provided us with a set of 1.2~million cross-matched stars.

@@

We warn that the SDSS comparison revealed a number of fields where the offsets
exceeded 0.05~mag (523 fields) or even 0.1~mag (18 fields).
Such sporadic outliers are consistent with the tails of a Gaussian
with mean $\approx0$ and $\sigma=0.03$.
Moreover, the SDSS calibration itself is not a perfect scale itself.

In future work, we hope to draw upon
the PanSTARRS survey \citep{Schlafly2012}
to further improve the accuracy of our calibration.
At the time of preparing this work data from PanSTARRS
had not been made public yet.

>>>>

Unlike the broadbands, 
standard-star photometry is not available in the literature 
for the H$\alpha$ passband
and hence there is no formally recognised flux scale 
for it.
We can specify here, however, 
that the integrated in-band energy flux for Vega 
in the IPHAS \ha\ filter 
is $1.52 \times 10^{-7}$ ergs\,cm$^{-2}$\,s$^{-1}$ 
at the top of the Earth's atmosphere,
which is the flux obtained by folding 
Vega's SED with the filter transmission curve 
(uncorrected for the atmosphere and detector quantum efficiency,
which would otherwise scale down the narrowband flux by 0.707).
This is 3.14 magnitudes less than the flux captured 
by the much broader $r$ band
which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}
Note that the above implies that the in-band flux corresponding to
zero magnitude is $1.56 \time 10^{-7}$ ergs\,cm$^{-2}$\,s$^{-1}$,
when the \ha\ magnitude for Vega is set by convention to 0.03.

@@

Unfortunately these schemes cannot be applied
directly to IPHAS
for two reasons. 
First, the survey was carried out 
in competitively-allocated observing time on a
common-user telescope, 
rendering the 2MASS approach 
of observing standards at a very high frequency
prohibitively expensive (it does not help that 
standard fields 
are very scarce within the Galactic Plane).
Second, IPHAS is not specified as a variability survey, with
the consequence that stars are not normally observed at more 
than one epoch,
unless they happen to fall within a narrow overlap region 
between two neighbouring field pairs.

Although the IPHAS data do contain a significant number of 
such inter-field repeat measurements,
we have found the information contained
in these regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is prone to systematics at the level of 1--2\%.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is more affected by a radial geometric distortion in the WFC,
which causes the pixel scale to increase as the edges 
are approached \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can combine, during the global calibration routine, to cause artificial zeropoint gradients 
across the survey unless controlled by
%when they are used to constrain a global calibration
%without 
other external constraints.

For these reasons, we have decided not to depend
on an ubercalibration-type scheme alone,
and have opted to involve an external reference survey
-- where available --
to bring the majority of our data onto a uniform calibration.
%This is explained next.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring 
%the vast majority - exaggerating a bit? 
most of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the original nightly calibrations of IPHAS,
for which we estimate the $1\sigma$-uncertainty to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

With the aim of bringing IPHAS to a similar accuracy of $\sim$3\%,
we used the APASS catalogue to identify and adjust the calibration of all IPHAS fields 
which showed a magnitude offset larger than 3\% against APASS.  Experience of re-running
the calibration and testing the results showed us that it was inadvisable to more
finely tune the match for IPHAS data obtained in what were generally the most photometric
nights.  To this end,
the $r$- and $i$-band detection tables of each IPHAS field
were cross-matched against the APASS DR7 catalogue 
using a maximum matching distance of 1~arcsec.
The magnitude range was limited to
$13<r_{\rm APASS}<16.5$ and $12.5<i_{\rm APASS}<16.0$
in order to avoid sources 
brighter than the IPHAS saturation limit on one hand, 
and to avoid sources near the faint detection limit of APASS 
on the other.

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the IPHAS and APASS broadband filters 
are very similar.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses Vega magnitudes.
Separate transformations were derived for sight-lines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
The variations at these different
sight-lines were found to be insignificant.
This is not surprising because heavily-reddened objects 
are naturally scarce at $r<16$.

Having transformed APASS magnitudes into the IPHAS system,
we then computed the median magnitude offset 
for each field which contained at least 30 cross-matched stars.
This was achieved for 48\% of our fields.
The mean offset was found to be
$0.014\pm0.104$~mag in $r$ and $0.007\pm0.108$~mag in $i$
(Table~\ref{tbl:offsets_before}).
A total of 4596 fields showed a median offset
exceeding $\pm$0.03~mag in either $r$ or $i$.

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of the fields in our data release.
Fortunately, these fields are located mainly 
in the early part of the Galactic Plane (Fig.~\ref{fig:apass_r}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme,
which minimises the magnitude offsets between stars
located in the overlap regions between neighbouring fields.
%Although we explained earlier that these overlap regions
%are prone to systematics, the use of APASS enabled us to
%keep the zeropoints of roughly half of the fields fixed,
%which avoids these systematics from combining to introduce
%artificial gradients across the survey.
%In the next section we will show this to be true by validating
%our calibration against SDSS.
An algorithm for achieving this minimisation
%of minimising the magnitude differences between overlapping frames 
has previously been described by~\citet{Glazebrook1994}.
In brief, there are two fundamental quantities to be
minimised between each pair of overlapping exposures, 
denoted by the indices $i$ and $j$. 
Firstly, the mean magnitude difference between stars in the overlap
region $\Delta_{ij}=\langle m_i-m_j\rangle=-\Delta_{ji}$ is a local
constraint. 
Secondly, to ensure the solution does not stray far 
from the existing calibration, 
the difference in zeropoints 
$\Delta\mathrm{ZP}_{ij}=-\Delta\mathrm{ZP}_{ji}$ 
between each pair of exposures must also be minimised.

@@

We enforce a strong external constraint
on the solution by keeping the zeropoint fixed 
for the 6\,756 fields which have already been compared
and calibrated against APASS.
We hereafter refer to these fields as \emph{anchors}.
It is asserted that the zeropoints $a_i$ of the anchor fields 
are known and not solved for.  However they do appear in the vector 
$b_j$ as constraints.  In addition to the APASS-based anchors, 
we selected 3\,273 additional anchor fields by hand
to provide additional constraints in regions not covered by APASS.
These extra anchors were deemed to have reliable zeropoints 
based on 
(i) the information contained in the observing logs,
(ii) the stability of the standard star zeropoints during the night, and
(iii) photometricity statistics provided by the Carlsberg Meridian Telescope,
which is located $\sim$500~m from the INT.

@@

We then turned to the uniform calibration of the \ha\ data.
It is not possible to re-calibrate the narrowband 
in the same way as the broadbands,
because the APASS survey does not offer \ha\ photometry.
We can reasonably assume however,
that the corrections required for $r$ and \ha\ are identical,
much of the time, because the IPHAS data-taking pattern ensured 
that a field's \ha\ and $r$-band exposures
were taken consecutively -- 
separated only by the $\sim$30~s read-out time.
Hence, we have corrected the \ha\ zeropoints 
by re-using the zeropoint adjustments that were derived for the $r$ band
in the earlier steps.
An exception was made for 3\,101 fields
for which our quality-control routines revealed
zeropoint variations which exceeded 0.03~mag 
between consecutive fields,
suggesting strongly non-photometric conditions.
In these cases, the \ha\ zeropoints of these fields
were adjusted by solving Eqn.~\ref{eqn:matrix}
rather than linking them directly to the $r$-band shift.

@@

Having re-calibrated all fields to the expected APASS accuracy of 3\%,
we turned to a different survey, SDSS Data Release 9 \citep{Ahn2012},
to validate the results.
SDSS DR9 includes several strips at low
Galactic latitudes,
providing data across 18\% of the fields in our data release.
We cross-matched the IPHAS fields against the subset of
objects marked as reliable stars in the SDSS catalogue\footnote{
We used the CasJobs facility located at http://skyserver.sdss3.org/CasJobs
to obtain photometry from the SDSS {\sc photoprimary} table 
with criteria {\sc type = star}, {\sc clean = 1} and {\sc score $>$ 0.7}.}
cross-matching in much the same way as for APASS,
with the difference of selecting from the  fainter magnitude ranges of 
$15<r_{\rm SDSS}<18.0$ and $14.5<i_{\rm SDSS}<17.5$.
This provided us with a set of 1.2~million cross-matched stars.

@@

We warn that the SDSS comparison revealed a number of fields where the offsets
exceeded 0.05~mag (523 fields) or even 0.1~mag (18 fields).
This pattern of outliers is consistent with the tails of a Gaussian
with mean $\approx0$ and $\sigma=0.03$.  Furthermore, it should not
be forgotten that both the SDSS and APASS calibrations are approximations to
perfection and will not be entirely free of anomalies.  Indeed as we worked, 
we noticed the occasional unsurprising examples of inconsistency between these 
two surveys.  
%, we hope to draw upon
%the PanSTARRS survey \citep{Schlafly2012}
%to further improve the accuracy of our calibration.
%At the time of preparing this work data from PanSTARRS
%had not been made public yet.
